---
title: "Assignment 2: Base R Graphics"
subtitle: "EPPS 6356 Data Visualization"
author: "Evandro M. S. Gomes"
date: today
format:
  html:
    code-fold: show
    toc: true
execute:
  warning: false
  message: false
---

## 1. Murrell, *R Graphics*, chapter 1

I ran the author's code for Figures 1.1 and 1.2 (<https://www.stat.auckland.ac.nz/~paul/RG3e/chapter1.html>) line by line. The `DEVICE:` comments note what changed on the device after each call. The code's comments pose no questions.

```{r}
#| label: fig-murrell
#| fig-height: 9
#| fig-width: 7
set.seed(1)
## Murrell, R Graphics (3rd ed.), chapter 1 - Figures 1.1 and 1.2
## Annotated walk-through: run ONE line at a time (Ctrl+Enter) and watch the
## Plots pane. "DEVICE:" notes say what changes after each call.

## ---- Figure 1.1: a simple scatterplot --------------------------------------
plot(pressure)
# DEVICE: new page; axes, box, labels and 19 points (temperature vs pressure).
text(150, 600, "Pressure (mm Hg)\nversus\nTemperature (Celsius)")
# DEVICE: nothing is erased; the text is added at DATA coordinates (x=150, y=600).
# "\n" breaks the label into three lines.

## ---- Figure 1.2: six standard plots on one page ---------------------------
par(mfrow = c(3, 2))
# DEVICE: nothing visible yet - the page is now split into a 3 x 2 grid,
# and each new high-level plot fills the next cell (row by row).

## (1) Scatterplot built piece by piece ------------------------------------
x  <- c(0.5, 2, 4, 8, 12, 16)
y1 <- c(1, 1.3, 1.9, 3.4, 3.9, 4.8)
y2 <- c(4, .8, .5, .45, .4, .3)
par(las = 1, mar = c(4, 4, 2, 4), cex = .7)
# DEVICE: nothing drawn; horizontal tick labels, wider right margin (for axis 4),
# smaller text.
plot.new()
# DEVICE: moves to cell 1 - an EMPTY plot region (no axes, no box).
plot.window(range(x), c(0, 6))
# DEVICE: still blank; only sets the coordinate system (x range, y 0-6).
lines(x, y1)
lines(x, y2)
# DEVICE: two lines appear, no axes yet.
points(x, y1, pch = 16, cex = 2)
# DEVICE: large filled circles on line 1.
points(x, y2, pch = 21, bg = "white", cex = 2)
# DEVICE: large open circles (white fill hides the line under them) on line 2.
par(col = "gray50", fg = "gray50", col.axis = "gray50")
# DEVICE: nothing yet; what is drawn next will be grey.
axis(1, at = seq(0, 16, 4))
axis(2, at = seq(0, 6, 2))
axis(4, at = seq(0, 6, 2))
# DEVICE: grey axes on bottom (1), left (2) and right (4) sides.
box(bty = "u")
# DEVICE: a U-shaped frame (left, bottom, right - no top line).
mtext("Travel Time (s)", side = 1, line = 2, cex = 0.8)
mtext("Responses per Travel", side = 2, line = 2, las = 0, cex = 0.8)
mtext("Responses per Second", side = 4, line = 2, las = 0, cex = 0.8)
# DEVICE: axis titles written in the MARGINS (line = distance from the plot).
text(4, 5, "Bird 131")
# DEVICE: label inside the plot, at data coordinates (4, 5).
par(mar = c(5.1, 4.1, 4.1, 2.1), col = "black", fg = "black", col.axis = "black")
# DEVICE: nothing; restores default margins and colours.

## (2) Histogram with density curve --------------------------------------
Y <- rnorm(50)
Y[Y < -3.5 | Y > 3.5] <- NA
x  <- seq(-3.5, 3.5, .1)
dn <- dnorm(x)
par(mar = c(4.5, 4.1, 3.1, 0))
hist(Y, breaks = seq(-3.5, 3.5), ylim = c(0, 0.5), col = "gray80", freq = FALSE)
# DEVICE: cell 2 - histogram of 50 random normals; freq = FALSE puts DENSITY on y.
lines(x, dnorm(x), lwd = 2)
# DEVICE: the theoretical normal curve is laid over the bars.
# (Run it again: the bars change, because rnorm() draws new random numbers.)
par(mar = c(5.1, 4.1, 4.1, 2.1))

## (3) Stacked barplot ---------------------------------------------------
par(mar = c(2, 3.1, 2, 2.1))
midpts <- barplot(VADeaths, col = gray(0.1 + seq(1, 9, 2)/11),
                  names = rep("", 4))
# DEVICE: cell 3 - four stacked bars, five grey shades, no group labels.
# barplot() also RETURNS the x-position of each bar's centre (midpts).
mtext(sub(" ", "\n", colnames(VADeaths)),
      at = midpts, side = 1, line = 0.5, cex = 0.5)
# DEVICE: group names written under each bar on two lines ("Rural\nMale").
text(rep(midpts, each = 5), apply(VADeaths, 2, cumsum) - VADeaths/2,
     VADeaths, col = rep(c("white", "black"), times = 3:2), cex = 0.8)
# DEVICE: each segment's value written in its middle; white text on the
# three dark segments, black on the two light ones.
par(mar = c(5.1, 4.1, 4.1, 2.1))

## (4) Side-by-side boxplots ---------------------------------------------
par(mar = c(3, 4.1, 2, 0))
boxplot(len ~ dose, data = ToothGrowth,
        boxwex = 0.25, at = 1:3 - 0.2,
        subset = supp == "VC", col = "white",
        xlab = "", ylab = "tooth length", ylim = c(0, 35))
# DEVICE: cell 4 - three narrow white boxes (Vitamin C), shifted LEFT by 0.2.
mtext("Vitamin C dose (mg)", side = 1, line = 2.5, cex = 0.8)
boxplot(len ~ dose, data = ToothGrowth, add = TRUE,
        boxwex = 0.25, at = 1:3 + 0.2,
        subset = supp == "OJ")
# DEVICE: add = TRUE draws on the SAME plot: grey boxes (orange juice)
# shifted RIGHT by 0.2, so each dose shows a pair.
legend(1.5, 9, c("Ascorbic acid", "Orange juice"),
       fill = c("white", "gray"), bty = "n")
# DEVICE: key at (1.5, 9), no border around it (bty = "n").
par(mar = c(5.1, 4.1, 4.1, 2.1))

## (5) Perspective (3-D) plot ------------------------------------------
x <- seq(-10, 10, length = 30)
y <- x
f <- function(x, y) { r <- sqrt(x^2 + y^2); 10 * sin(r)/r }
z <- outer(x, y, f)
z[is.na(z)] <- 1
# (sin(0)/0 is NaN at the centre; it is replaced by the limit value.)
par(mar = c(0, 0.5, 0, 0), lwd = 0.5)
persp(x, y, z, theta = 30, phi = 30, expand = 0.5)
# DEVICE: cell 5 - the "sombrero" surface; theta turns it left/right,
# phi tilts it up/down, expand squashes the height.
par(mar = c(5.1, 4.1, 4.1, 2.1), lwd = 1)

## (6) Pie chart ---------------------------------------------------------
par(mar = c(0, 2, 1, 2), xpd = FALSE, cex = 0.5)
pie.sales <- c(0.12, 0.3, 0.26, 0.16, 0.04, 0.12)
names(pie.sales) <- c("Blueberry", "Cherry", "Apple",
                      "Boston Cream", "Other", "Vanilla")
# names() draws nothing: it only attaches labels to the vector,
# and pie() then uses those names as slice labels.
pie(pie.sales, col = gray(seq(0.3, 1.0, length = 6)))
# DEVICE: cell 6 - six grey slices, labelled from names(pie.sales).
par(mar = c(5.1, 4.1, 4.1, 2.1), xpd = FALSE, cex = 1, mfrow = c(1, 1))
# DEVICE: back to one plot per page for what follows.
```

## 2. Data: the Happy Planet Index

Source: Abdallah, S. & Barberà Mas, E. (2026). *The Happy Planet Index 2026 Data File*. Hot or Cool Institute. <https://happyplanetindex.org>.

**Year: 2025.** I chose 2025 because it is the most recent year in the release. It covers 134 countries.

`names()` is used only to give the columns short names; it is not a graphics function.

```{r}
#| label: data
library(readxl)
## Data: HPI 2026 data file, sheet "1. All countries" (year selected: 2025).
## The first 8 rows are titles; the column names are in row 9.
hpi <- read_excel("Data/HPI_public_dataset.xlsx",
                  sheet = "1. All countries", skip = 8)
hpi <- hpi[, 1:12]
names(hpi) <- c("rank", "country", "iso", "blank", "continent", "pop",
                "lifeexp", "satisf", "footprint", "hpi", "threshold", "gdp")
hpi <- hpi[!is.na(hpi$country) & !is.na(hpi$hpi), ]
hpi$continent <- factor(hpi$continent, levels = 1:8,
                        labels = c("Latin America", "N. America & Oceania",
                                   "Western Europe", "Middle East & N. Africa",
                                   "Sub-Saharan Africa", "South Asia",
                                   "E. Europe & Central Asia", "East Asia"))
nrow(hpi)
```

## 3. The plotting functions

### a. Set up the canvas

#### `par()`

`par()` sets graphical parameters for the plots that follow; here `mfrow = c(1, 2)` puts two plots side by side.

```{r}
par(mfrow = c(1, 2))
plot(hpi$satisf, hpi$hpi, xlab = "Life satisfaction", ylab = "HPI")
plot(hpi$footprint, hpi$hpi, xlab = "Ecological footprint", ylab = "HPI")
par(mfrow = c(1, 1))
```

#### `box()`

`box()` draws the frame around the plot region; `bty = "l"` keeps only the left and bottom sides.

```{r}
plot(hpi$lifeexp, hpi$hpi, axes = FALSE,
     xlab = "Life expectancy", ylab = "HPI")
axis(1)
axis(2)
box(bty = "l")
```

#### `axis()`

`axis()` draws one axis and controls where its tick marks go and how they are labelled.

```{r}
plot(hpi$gdp, hpi$hpi, log = "x", axes = FALSE,
     xlab = "GDP per capita (log scale)", ylab = "HPI")
axis(1, at = c(1000, 10000, 100000), labels = c("$1k", "$10k", "$100k"))
axis(2)
box()
```

### b. Draw whole plots

#### `hist()`

`hist()` divides a numeric variable into bins and draws how many observations fall in each.

```{r}
hist(hpi$hpi, col = "gray80", xlab = "HPI", main = "HPI scores, 2025")
```

#### `boxplot()`

`boxplot()` shows the median, quartiles and outliers of a variable for each group.

```{r}
par(mar = c(4, 11, 2, 1))
boxplot(hpi ~ continent, data = hpi, horizontal = TRUE, las = 1,
        xlab = "HPI", ylab = "")
par(mar = c(5.1, 4.1, 4.1, 2.1))
```

#### `pie()`

`pie()` divides a circle into slices proportional to a set of values.

```{r}
pie(table(hpi$continent), main = "Countries by continent")
```

#### `persp()`

`persp()` draws a 3-D surface seen from an angle set by `theta` and `phi`; here the density of countries by life expectancy and footprint.

```{r}
dens <- MASS::kde2d(hpi$lifeexp, hpi$footprint)
persp(dens$x, dens$y, dens$z, theta = -30, phi = 30,
      xlab = "Life expectancy", ylab = "Footprint", zlab = "Density")
```

### c. Add to an existing plot

#### `points()`

`points()` adds points to an existing plot; here Brazil in red.

```{r}
plot(hpi$footprint, hpi$hpi, col = "gray",
     xlab = "Ecological footprint", ylab = "HPI")
br <- hpi$country == "Brazil"
points(hpi$footprint[br], hpi$hpi[br], pch = 16, col = "red", cex = 2)
```

#### `lines()`

`lines()` adds connected lines to an existing plot; here a lowess trend.

```{r}
plot(hpi$footprint, hpi$hpi, xlab = "Ecological footprint", ylab = "HPI")
lines(lowess(hpi$footprint, hpi$hpi), lwd = 2)
```

### d. Annotate

#### `text()`

`text()` writes text inside the plot at given coordinates; here two country names.

```{r}
plot(hpi$satisf, hpi$lifeexp, xlab = "Life satisfaction",
     ylab = "Life expectancy")
sel <- hpi$country %in% c("Brazil", "Costa Rica")
text(hpi$satisf[sel], hpi$lifeexp[sel], hpi$country[sel], pos = 2)
```

#### `mtext()`

`mtext()` writes text in the margins of the plot; here the data source.

```{r}
plot(hpi$satisf, hpi$hpi, xlab = "Life satisfaction", ylab = "HPI")
mtext("Source: Happy Planet Index 2026", side = 1, line = 4, adj = 0)
```

#### `legend()`

`legend()` adds a key that explains colours or symbols; here the continents.

```{r}
plot(hpi$footprint, hpi$hpi, col = hpi$continent, pch = 16,
     xlab = "Ecological footprint", ylab = "HPI")
legend("topright", legend = levels(hpi$continent), col = 1:8, pch = 16,
       cex = 0.7)
```

## Reference

Murrell, Paul. 2019. *R Graphics*. 3rd ed. Chapman & Hall/CRC.



























