---
title: "Assignment 3: Anatomy of a Chart"
subtitle: "EPPS 6356 Data Visualization"
author: "Evandro M. S. Gomes"
date: today
format:
  html:
    code-fold: show
    toc: true
execute:
  warning: false
  message: false
---

## 1. Document a chart

I reran `murrell01.R` in full. Then I chose the **"Bird 131" scatterplot**.

```{r}
#| label: fig-bird131
#| fig-cap: "Bird 131 (Murrell, R Graphics)."
#| fig-width: 7
#| fig-height: 5

# Data: travel time (x) and two response measures (y1, y2).
x  <- c(0.5, 2, 4, 8, 12, 16)
y1 <- c(1, 1.3, 1.9, 3.4, 3.9, 4.8)
y2 <- c(4, .8, .5, .45, .4, .3)

# Horizontal axis labels, margins (bottom, left, top, right) and smaller text.
# The right margin is bigger to fit a second y axis.
par(las = 1, mar = c(4, 4, 2, 4), cex = .7)

# Opens an empty plot. Nothing is drawn yet.
plot.new()

# Sets the x and y ranges. Nothing is drawn yet.
plot.window(range(x), c(0, 6))

# Draws the line for y1.
lines(x, y1)

# Draws the line for y2.
lines(x, y2)

# Adds black filled points to y1 (pch = 16), twice the size (cex = 2).
points(x, y1, pch = 16, cex = 2)

# Adds open points to y2 (pch = 21) with a white inside (bg).
# Filled and open points show the two series without a legend.
points(x, y2, pch = 21, bg = "white", cex = 2)

# From here on, axes and labels are gray, so the data stands out.
par(col = "gray50", fg = "gray50", col.axis = "gray50")

# x axis at the bottom (side 1).
axis(1, at = seq(0, 16, 4))

# y axis on the left (side 2).
axis(2, at = seq(0, 6, 2))

# y axis on the right (side 4).
axis(4, at = seq(0, 6, 2))

# Frame around the plot, open at the top ("u" shape).
box(bty = "u")

# x axis title, in the bottom margin.
mtext("Travel Time (s)", side = 1, line = 2, cex = 0.8)

# Left y axis title; las = 0 makes it vertical.
mtext("Responses per Travel", side = 2, line = 2, las = 0, cex = 0.8)

# Right y axis title, also vertical.
mtext("Responses per Second", side = 4, line = 2, las = 0, cex = 0.8)

# Label inside the plot, at x = 4 and y = 5.
text(4, 5, "Bird 131")

# Resets margins and colors to the default.
par(mar = c(5.1, 4.1, 4.1, 2.1), col = "black", fg = "black",
    col.axis = "black")

```

## 2. Anscombe's quartet

I reran `anscombe01.R` in full.

### Regression models

```{r}
# Fit one linear regression for each of the four sets.
lm1 <- lm(y1 ~ x1, data = anscombe)
lm2 <- lm(y2 ~ x2, data = anscombe)
lm3 <- lm(y3 ~ x3, data = anscombe)
lm4 <- lm(y4 ~ x4, data = anscombe)
```

The numbers below come from `summary(lm1)` to `summary(lm4)`.

|                         | Set 1  | Set 2 | Set 3  | Set 4  |
|-------------------------|--------|-------|--------|--------|
| Intercept               | 3.0001 | 3.001 | 3.0025 | 3.0017 |
| Slope                   | 0.5001 | 0.500 | 0.4997 | 0.4999 |
| R²                      | 0.6665 | 0.6662| 0.6663 | 0.6667 |
| Residual standard error | 1.237  | 1.237 | 1.236  | 1.236  |

The four models are almost identical: y ≈ 3 + 0.5x, with the same R² and the same error.

### Four ways to draw them

```{r}
#| label: fig-anscombe
#| fig-cap: "Anscombe's quartet: same regression line, four different patterns."
#| fig-width: 8
#| fig-height: 8

# Four plots on one page: 2 rows, 2 columns.
par(mfrow = c(2, 2))

# Set 1: blue circles, solid line.
plot(anscombe$x1, anscombe$y1, col = "blue", pch = 16,
     xlim = c(3, 19), ylim = c(3, 13), xlab = "x1", ylab = "y1")
abline(lm1, col = "gray40", lty = 1, lwd = 2)

# Set 2: green triangles, dashed line.
plot(anscombe$x2, anscombe$y2, col = "forestgreen", pch = 17,
     xlim = c(3, 19), ylim = c(3, 13), xlab = "x2", ylab = "y2")
abline(lm2, col = "gray40", lty = 2, lwd = 2)

# Set 3: yellow squares, dotted line.
plot(anscombe$x3, anscombe$y3, col = "gold", pch = 15,
     xlim = c(3, 19), ylim = c(3, 13), xlab = "x3", ylab = "y3")
abline(lm3, col = "gray40", lty = 3, lwd = 2)

# Set 4: red diamonds, dot-dash line.
plot(anscombe$x4, anscombe$y4, col = "red", pch = 18,
     xlim = c(3, 19), ylim = c(3, 13), xlab = "x4", ylab = "y4")
abline(lm4, col = "gray40", lty = 4, lwd = 2)

# Back to one plot per page.
par(mfrow = c(1, 1))
```

**Comparing the drawings.** 

Professor's version: 
- same style in all four plots: orange points and a blue line.

My version:
- each set has its own color, its own symbol, and its own line type.

**What the summary statistics hide.** 
The four regressions have the same intercept (3), slope (0.5), R² (0.67) and residual standard error (1.24). The plots show:

- **Set 1:** a normal linear relationship, with points around the line.
- **Set 2:** a curve. A straight line is the wrong model.
- **Set 3:** an almost perfect line, but one outlier pulls the regression line up.
- **Set 4:** almost all points have x = 8. One single point at x = 19 creates the whole line.

Same numbers, four stories. Always plot the data before trusting the statistics.

## 3. Finetune without leaving base R

I used Anscombe's Set 1 with a serif font (Garamond), non-default colors, and my own plotting character (the letter "E").

```{r}
#| label: fig-tuned
#| fig-width: 7
#| fig-height: 6

lm1 <- lm(y1 ~ x1, data = anscombe)

windowsFonts(Garamond = windowsFont("Garamond"))
par(mfrow = c(1, 1), family = "Garamond")

plot(anscombe$x1, anscombe$y1, xlab = "x1", ylab = "y1",
     main = "Anscombe's Set 1",
     col = "darkblue", pch = "E", cex = 1.3,
     col.main = "darkblue", col.lab = "gray30", col.axis = "gray30")
abline(lm1, col = "firebrick", lwd = 2)
```
## 4. Now do it in ggplot2

I reproduced the tuned chart from Part 3 with ggplot2.

```{r}
#| label: fig-ggplot
#| fig-width: 7
#| fig-height: 6

library(tidyverse)
windowsFonts(Garamond = windowsFont("Garamond"))
lm1 <- lm(y1 ~ x1, data = anscombe)

ggplot(anscombe, aes(x = x1, y = y1)) +
  geom_text(label = "E", family = "Garamond", colour = "darkblue", size = 6) +
  geom_abline(intercept = coef(lm1)[1], slope = coef(lm1)[2],
              colour = "firebrick", linewidth = 1) +
  scale_x_continuous(breaks = seq(4, 14, 2)) +
  scale_y_continuous(breaks = 4:11) +
  labs(title = "Anscombe's Set 1", x = "x1", y = "y1") +
  theme_bw() +
  theme(text = element_text(family = "Garamond"),
        plot.title = element_text(colour = "darkblue", face = "bold", hjust = 0.5),
        axis.title = element_text(colour = "gray30"),
        axis.text = element_text(colour = "gray30"),
        panel.grid = element_blank())
```

**Became `theme()` arguments:** the font (`family` in `par()` became `text = element_text(family = ...)`), the title color (`col.main` became `plot.title`), the axis label color (`col.lab` became `axis.title`) and the axis number color (`col.axis` became `axis.text`).

**Became `scale_*()`:** the tick marks on the axes, which base R chose by itself, became `scale_x_continuous(breaks = ...)` and `scale_y_continuous(breaks = ...)`.

**No equivalent:** `windowsFonts()` and `par()` have no ggplot2 version. Also, `pch = "E"` had no direct match: `geom_point(shape = "E")` ignores the theme font, so I used `geom_text()`.

